Community-Centric Innovation

### 2014-03-01 paper.tex
final version

\begin{abstract}
%\boldmath
The Internet is profoundly transforming the economic and social structures of the 21st century.
The ability to instantaneously share data across the globe enables new ways of collaboration and new production models.
The individuals can join loosely organized working groups and communities and contribute to the evolution and progress of the humankind.
By choosing which communities they join and which communities they contribute to, the users can also be participants and producers.
%Furthermore, this model enables a new way of democracy in which the people vote not only with their ballots, but also with their daily jobs and contributions.
In this work we describe examples of community-centric innovation in which the users play a central role in the development of the project.
The emphasis is placed in community networks and networked cryptocurrencies.
These technologies are good examples in the sense that they empower the communities that use and develop them.
These are initiatives by and for the people.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
 bottom-up broadband, community networks
%Slotted Aloha, game theory, contention control, media access control.
\end{IEEEkeywords}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{he} growth of the Internet has dramatically accelerated the path of technological evolution.
The amount of information and knowledge available on the Net is increasing.
Importantly,  the number of people that can access that knowledge is increasing as well.
The Internet is more than an incremental evolution that brings faster progress.
It is a disruptive innovation that is completely transforming the way we organize ourselves and relate to each other.

The fact that the information can be shared with little restrictions in terms of space, time and cost, makes it possible for people to organize in horizontal structures.
In this horizontal models, also known as peer-to-peer models, all the participants can access to the information required to contribute to the decision-making process.
By removing the barriers for participation, the working force of peer-to-peer projects is reinforced.
The benefit is not only the number of participants, but also the variety.
The heterogeneous background of the participants makes it possible to consider a wider space of solutions and the extensive scrutiny throughout the discussion and decision-making stage reduces the chances of mistakes.

@@

Compared to commercial Internet, Community Networks are more heterogeneous and decentralized.
As a consequence they have properties that complement those of commercial networks.
In case of natural disaster, having alternative resilient networks to coordinate rescue and humanitarian help can save lives.

@@

All participants are invited to contribute, and contributions are evaluated in an informal process for merit or weaknesses.
The participants can progressively build up a reputation by making high-quality contributions and helping others to accomplish their goals.
The organization progressively evolves towards a meritocratic systems in which those participants that consistently contribute to the value of the community earn the respect of their peers and emerge as natural leaders.
In any case, this is not a leadership which is attached to the person. 
The leadership is attached to the contributions.

@@

Peer-to-peer networks are extremely resilient.
There is no single point of failure or central entity that can be attacked to take down the network.
As there are thousands of nodes spread across the network, the currency can survive as long as the underlying data network survives.
The risks of blockades or political interference are greatly mitigated.
Many Internet organizations such as Wikileaks and Wordpress accept donations and payments in cryptocurrency.
In some cases the reason to bypass government or paying networks blockades.
In others, the motivation is to operate in countries with lacking payment networks.

It is in areas lacking effective financial and banking systems where cryptocurrencies have a potential for greater impact.
Just like some areas leapfrogged from no telecommunication services to wireless handheld Internet enabled devices, it is possible that these same areas, that currently use mobile calls minutes as a currency, transition directly to the latest currency technology which is cryptocurrency.
Many countries suffer from chronic hyperinflation problems that could be solved by moving to a cryptocurrency in which the inflation rate (if any) can be chosen by the community and embedded in the protocol.

@@

The use of networked criptocurrencies evidences that flows of money are nothing more than flows of information.
Money transactions are simply communications among the involved parties.
The freedom to use different currencies is directly related to the freedom to communicate, which is popularly referred to as free speech.
The flows of money within a community are communications flows that strengthen the relation between the members of the community and, therefore, the community as a whole.

@@

Throughout human history, organizational models have changed, transformed and evolved.
As new technologies develop, old models are progressively replaced by others that are more efficient in the context of the new technologies.
The invention of information processing and transmission tools multiplied the potential of community-centric models.
Initiatives embracing the community-centric model have an advantage compared to those attaching to old hierarchical approaches.

@@

Community-centric innovation such as Community Networks and Networked Cryptocurrencies have attracted the attention of the research community.
In this section we detail three European projects that are related to community-centric innovation.

@@

\subsection{Confine}
The Confine project \cite{braem2013crc} emphasizes the links between research and community networks.
The goal is to make it easier for researchers to run experiments in community networks.
It provides the tools for automatizing and monitoring the experiments, as well as the hardware that runs the experiments.
These hardware comprises Confine nodes which are distributed across the community network.

@@

Despite being more time-consuming and less reliable for the user, they are still a valid alternative for some people.
The possibility of controlling the network and being participants instead of consumers are some of the reasons for joining a community network.
Another reason is that the idea of freedom is embedded in community networks.
Community networks can resist authoritarian censorship or Internet blackouts ordered by governments.
They are also extremely useful in case of a natural disaster when centralized infrastructures may collapse.

Networked cryptocurrencies are another example of community-centric innovation.
These currencies are developed by the communities as a result of open debate and discussion.
There are many similitudes with community networks.
The users of networked cryptocurrencies can contribute one node to the network to help in maintaining security.
The members of the community can also participate in many other ways offering products and services and using the cryptocurrency in their transactions.Networked cryptocurrencies are also resilient to government interference or blockade.
They preserve the freedom of the user to spend the money in whatever she believes appropriate.

Community-centric innovation offers today tools that greatly empower the communities.
This opportunity can be captured to further evolve the existing tools and develop new ones that can facilitate the freedom and well-being of the population.

>>>>

\begin{abstract}
%\boldmath
The Internet is profoundly transforming the economic and social structures of the 21st century.
The ability to instantaneously share data across the globe enables new ways of collaboration and new production models.
The individuals can join loosely organized working groups and communities and contribute to the evolution and progress of the humankind.
By choosing which communities they join and which communities they contribute to, the users can also be participants and producers.
%Furthermore, this model enables a new way of democracy in which the people vote not only with their ballots, but also with their daily jobs and contributions.
In this work we describe examples of community-centric innovation in which the users play a central role.
The emphasis is placed in community networks and networked cryptocurrencies.
These technologies are good examples in the sense that they empower the communities that use and develop them.
These are initiatives by and for the people.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
 community networks, cryptocurrencies, peer-to-peer production models
%Slotted Aloha, game theory, contention control, media access control.
\end{IEEEkeywords}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{he} growth of the Internet has dramatically accelerated the pace of technological evolution.
The amount of information and knowledge available on the Net is increasing.
Importantly,  the number of people that can access that knowledge is increasing as well.
The Internet is more than an incremental evolution that brings faster progress.
It is a disruptive innovation that is completely transforming the way we organize ourselves and relate to each other.

The fact that the information can be shared with little restrictions in terms of space, time and cost, makes it possible for people to organize in horizontal structures.
In these horizontal models, also known as peer-to-peer models, all the participants can access to the relevant information to contribute to the decision-making process.
By removing the barriers for participation, the working force of peer-to-peer projects is reinforced.
The benefit is not only the number of participants, but also the variety.
The heterogeneous background of the participants makes it possible to consider a wider space of solutions and the extensive scrutiny throughout the discussion and decision-making stage reduces the chances of mistakes.

@@

Compared to commercial Internet, community networks are more heterogeneous and decentralized.
As a consequence they have properties that complement those of commercial networks.
In case of natural disaster, having alternative resilient networks to coordinate rescue and humanitarian help can save lives.

@@

All participants are invited to contribute, and contributions are evaluated in an informal process for merit or weaknesses.
The participants can progressively build up a reputation by making high-quality contributions and helping others to accomplish their goals.
The organization progressively evolves towards a meritocratic system in which those participants that consistently contribute to the community earn the respect of their peers and emerge as natural leaders.
In any case, this is not a leadership which is attached to the person. 
The leadership is attached to the contributions.

@@

Peer-to-peer networks are extremely resilient.
There is no single point of failure or central entity that can be attacked to take down the network.
As there are thousands of nodes spread across the network, the currency can survive as long as the underlying data network survives.
The risks of blockades or political interference are greatly mitigated.
Many Internet organizations such as Wikileaks and Wordpress accept donations and payments in cryptocurrency.
In some cases, the reason is to bypass financial blockades.
In others, the motivation is to operate in countries with lacking payment networks.

It is in areas lacking effective financial and banking systems where cryptocurrencies have a potential for greater impact.
Some areas leapfrogged from no telecommunication services to wireless handheld Internet enabled devices.
It is possible that these same areas, that currently use mobile calls minutes as a currency, transition directly to the latest currency technology which is cryptocurrency.
Many countries suffer from chronic hyperinflation problems that could be solved by moving to a cryptocurrency in which the inflation rate (if any) can be chosen by the community and embedded in the protocol.

@@

The use of networked criptocurrencies evidences that flows of money are nothing more than flows of information.
Money transactions are simply communications among the involved parties.
The freedom to use different currencies is directly related to the freedom to communicate, which is popularly referred to as free speech.
The flows of money within a community are communication flows that strengthen the bonds between the members of the community and, therefore, the community as a whole.

@@

Throughout human history, organizational models have changed, transformed and evolved.
As new technologies develop, old models are progressively replaced by others that are more efficient in the context of those new technologies.
The invention of information processing and transmission tools multiplied the potential of community-centric models.
Initiatives embracing the community-centric model have an advantage compared to those attaching to old hierarchical approaches.

@@

Community-centric innovations such as community networks and networked cryptocurrencies have attracted the attention of the research community.
In this section we detail three European projects that are related to community-centric innovation.

@@

\subsection{Confine}
The Confine project \cite{braem2013crc} emphasizes the links between research and community networks.
The goal is to make it easier for researchers to run experiments in community networks.
It provides the tools for automatizing and monitoring the experiments, as well as the hardware that runs the experiments.
This hardware comprises Confine nodes which are distributed across the community network.

@@

Despite being more time-consuming and less reliable for the user, they are still a valid alternative for some people.
The possibility of controlling the network and being participants instead of consumers are some of the reasons for joining a community network.
Another reason is that the idea of freedom is embedded in community networks.
Community networks can resist authoritarian censorship, surveillance or Internet blackouts ordered by governments.
They are also extremely useful in case of a natural disaster when centralized infrastructures may collapse.

Networked cryptocurrencies are another example of community-centric innovation.
These currencies are developed by the communities as a result of open debate and discussion.
There are many similitudes with community networks.
The users of networked cryptocurrencies can contribute one node to the network to help in maintaining security.
The members of the community can also participate in many other ways offering products and services and using the cryptocurrency in their transactions.
Networked cryptocurrencies are also resilient to government interference or blockade.
They preserve the freedom of the user to spend the money in whatever she believes appropriate.

Community-centric innovation offers today tools that greatly empower the communities.
New production models arise in which people can collaborate in a peer-to-peer fashion to effectively produce commons goods.
This opportunity can be seized to further evolve the existing tools and develop new ones that can facilitate the freedom and well-being of the population.
